Statistical framework for whole-genome bisulfite-sequencing data

### 2014-12-07 Rmarkdown/wgbs_statistical_framework.Rmd
Genome-wide beta distributions

Restricting our attention to CpGs, __FIGURES__ show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively. __FIGURES__ show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.

What is immediately clear is that these distributions are bimodal. The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}.

The only samples with significant intermediate methylation are either somatic cell lines --- __LIST SAMPLES__ --- or cancer cells lines or tissue --- __LIST SAMPLES__. __EXPLANATION FOR CANCER__. __EPISCOPE DATA SHOWS ITS NOT SOMATIC CELLS BUT CELL LINES THAT GET INTERMEDIATE METHYLATION__. __HYPOTHESIS FOR SOMATIC CELL LINES__

- Most samples are hypermethylated

__ALL THE GENOME-WIDE FIGURES__

__CGI FIGURES__

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. Most CpGs are highly unmethylated or highly methylated, with the buffy coat samples being the most extreme example of this bimodality. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose.}
\label{fig:EPISCOPE_beta_distribution}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution_CGI_stratified.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Most CpGs outside of CpG islands are methylated whereas most CpGs within CpG islands are unmethylated. Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: \em{BUF} = buffy coat; \em{SA} = subcutaneous adipose; \em{VA} = visceral adipose; \em{VAT} = adipocytes derived from visceral adipose; \em{CGI} = CpG island.}
\label{fig:EPISCOPE_beta_distribution_CGI_stratified}
\end{figure}

>>>>

Restricting our attention to CpGs, Figures \ref{fig:EPISCOPE_plotBetaDensity, fig:Lister_plotBetaDensity, fig:Seisenberger_plotBetaDensity, fig:Zilelr_plotBetaDensity} show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively. Figures \ref{fig:EPISCOPE_plotBetaFreqPoly, fig:Lister_plotBetaFreqPoly, fig:Seisenberger_plotBetaFreqPoly, fig:Zilelr_plotBetaFreqPoly} show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.

What is immediately clear is that these distributions are bimodal. The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.

Almost all the samples with significant intermediate methylation are either somatic cell lines --- _ADS_, _ADS-adipose_, _FF_, _IMR90\_r1_ and _IMR90\_r2_ from the Lister dataset; _SRR949208_ and _SRR949209_ from the Ziller dataset --- or cancer cells lines and tissue --- _SRR949206_, _SRR949207_, _SRR949210_, _SRR949211_, _SRR949212_, _SRR949213_ and _SRR949215_ from the Ziller dataset. Aside from the aforementioned _E16.5\_male\_1_, the _E6.5\_epiblast\_1_ sample from the Seisenberger dataset also displays greater levels of intermediate methylation. This sample was also created by pooling DNA from multiple mice, which may explain the extra variability in the $\beta$-value distribution.

It has been previously observed that cancer samples have highly variable DNA methylation \citep{Hansen:2011gu}, which, combined with the possibility of multiple sub-clones, explains these intermediate $\beta$-values in the cancer samples. The explanation for the somatic cell lines is less clear. Notably, all of the 12 EPISCOPE samples, which are somatic tissue samples rather than cell lines, have relatively low levels of intermediate methylation. This raises the question as to whether the widespread partial methylation observed in the somatic samples from the Lister dataset is in fact a feature of somatic cell lines rather than somatic cells _per se_. Naively, a cell line is a 'pure' cell population, however, the DNA methylation data clearly reveal widespread differences between cells in these populations.

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_plotBetaDensity}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:EPISCOPE_plotBetaFreqPoly}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Lister_plotBetaFreqPoly}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Seisenberger_plotBetaFreqPoly}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Ziller_plotBetaFreqPoly}
\end{figure}

__TODO: DESCRIBE CGI FIGURES__

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: \em{CGI} = CpG island.}
\label{fig:EPISCOPE_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: \em{CGI} = CpG island.}
\label{fig:Lister_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: \em{CGI} = CpG island.}
\label{fig:Seisenberger_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: \em{CGI} = CpG island.}
\label{fig:Ziller_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status. Abbreviations: \em{CGI} = CpG island.}
\label{fig:EPISCOPE_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status. Abbreviations: \em{CGI} = CpG island.}
\label{fig:Lister_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status. Abbreviations: \em{CGI} = CpG island.}
\label{fig:Seisenberger_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/Ziller_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Zilelr data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status. Abbreviations: \em{CGI} = CpG island.}
\label{fig:Ziller_plotBetaFreqPoly_CGI}
\end{figure}
